---
title: Analyzing business ownership demographics from Annual Business Survey
author: ''
date: '2020-06-14'
slug: analyzing-business-ownership-demographics-from-annual-business-survey
categories: []
tags:
  - R
subtitle: ''
summary: 'using 2018 census ABS data'
authors: []
lastmod: '2020-06-14T11:36:49-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Last month, U.S. Census Bureau released its first ever Annual Business Survey [ABS](https://www.census.gov/programs-surveys/abs.html), covering statistics of nonfarm employer businesses in 2017.
ABS provides business and business owner characteristics such as firms, receipts, payroll and employment by demographic characteristics such as gender, race, ethnicity and veteran status, at regional level. 

Today I'm going to demonstrate how to fetch the ABS data using [Census API](https://www.census.gov/data/developers/data-sets.html), and explore the demographics of business ownership across cities, counties, and metropolitan areas in the U.S.

## Fetch ABS data using Census API
There are many great packages and tutorials that walk you through the process of using Census API, so I won't go into the details. My favorite one is [tidycensus](https://walker-data.com/tidycensus/), but unfortunately it's available for American Community Survey and Decennial Survey only, so I'm going to use [censusapi](https://hrecht.github.io/censusapi/index.html) today, which offers a more flexible query structure.

Let's first take a look at the [ABS API](https://www.census.gov/programs-surveys/abs/data/api.html) endpoints. Navigate to its [variables](https://api.census.gov/data/2017/abscs/variables.html) page, it lists all the variables you can query from this API. 

Here are the variables I decide to include:
```{r variable list, message=F}
library(tidyverse)
abs_var <- c("GEO_ID", "NAICS2017", "NAICS2017_LABEL", "SEX", "SEX_LABEL", "ETH_GROUP", "ETH_GROUP_LABEL", "RACE_GROUP", "RACE_GROUP_LABEL", "VET_GROUP", "VET_GROUP_LABEL", "FIRMPDEMP", "RCPPDEMP", "EMP", "PAYANN")
```

Now I'm going to test my query with national data first. 

```{r fetch data}

censusapi::getCensus(
  name = "abscs",
  vintage = 2017,
  vars = abs_var,
  region = "us:*",
  key = Sys.getenv("CENSUS_API_KEY")
) %>% head()

```


Now let's get the data for every metro areas and counties. 

```{r get county and metro data}

cbsa_abs_naics <- getCensus(
  name = "abscs",
  vintage = 2017,
  vars = abs_var,
  region = "metropolitan statistical area/micropolitan statistical area:*",
  key = Sys.getenv("CENSUS_API_KEY")
) %>% 
  mutate(GEOID = str_sub(GEO_ID, -5,-1)) 

stco_abs_naics <- getCensus(
  name = "abscs",
  vintage = 2017,
  vars = abs_var,
  region = "county:*",
  key = Sys.getenv("CENSUS_API_KEY")
) %>% 
  mutate(GEOID = str_sub(GEO_ID, -5,-1))

```


## Exploratory Data Analysis
How does business ownership demographics look like in Detroit and Birmingham? 

First, let's specify the [GEOIDs](https://www.census.gov/programs-surveys/geography/guidance/geo-identifiers.html) of the place of interests. If you are not familiar with census geographies, MCDC has a very friendly GEOID [lookup](https://census.missouri.edu/geocodes/) tool.  

```{r target regions}

target_cbsa <- c("19820","13820")   # cbsa codes for Detroit-Warren-Dearborn, MI and Birmingham-Hoover, AL

```

Now we can find everything about these two metro areas by filtering the raw data. Use `count()` to get the codebook of `NAICS2017`, `SEX`, `RACE_GROUP`, `ETH_GROUP`, `VET_GROUP`. For example:

```{r codebook}
cbsa_abs_naics %>% 
  count(SEX, SEX_LABEL)
```


Let's take a look at female ownership. So we will get data for female (`002`), and all businesses that could be classified by gender (`096`).

```{r filter metro data}
cbsa_result <- cbsa_abs_naics %>% 
  filter(GEOID %in% target_cbsa) %>% 
  filter(SEX %in% c("002","096")) %>% 
  filter(NAICS2017 == "72") %>% 
  select(GEOID, contains("_LABEL"), FIRMPDEMP, EMP, RCPPDEMP, PAYANN) 

cbsa_result

```

I'm going to reshape the data to calculate number of firms (`FIRMPDEMP`), employment (`EMP`), receipts (`RCPPDEMP`), and payroll (`PAYANN`) of female-owed businesses in these two metro areas. 

```{r reshape and plot}
cbsa_result %>% 
  pivot_longer(FIRMPDEMP:PAYANN) %>%
  mutate(value = as.numeric(value)) %>% 
  select(GEOID, NAICS2017_LABEL, SEX_LABEL, name, value) %>% 
  pivot_wider(names_from = "SEX_LABEL", values_from = "value") %>% 
  mutate(pct_female = Female/Classifiable) %>% 
  
  ggplot(aes(x = name, y = pct_female, fill = name, label = scales::percent(pct_female, accuracy = 0.1)))+
  geom_col()+
  geom_text(vjust = -0.5)+
  facet_wrap(~GEOID)+
  scale_y_continuous(labels = scales::percent, "share of female owned businesses")+
  theme_classic()
```


We can see that in the accommodation sector, female business ownership is similar in Birmingham and Detroit (17.3% and 18.7%), but female-owned firms are much smaller in size in Birmingham. They only account for 11.5% of total employment, 12% of payroll, and 10.6% of total sales and receipts. 

## Scale it up
Let's write a function to easily adapt this script to explore explore other demographic groups in other regions. 

```{r function}

summarise_abs <- function(df, target, col, col_var, NAICS = "00"){
  col <- enquo(col)
  col_LABEL <- paste0((rlang::as_label(col)),"_LABEL")
    
df %>% 
  filter(GEOID %in% target) %>% 
  filter(!!col %in% col_var) %>% 
  filter(NAICS2017 == NAICS) %>%
    
  select(GEOID, contains("_LABEL"), FIRMPDEMP, EMP, RCPPDEMP, PAYANN) %>% 
  pivot_longer(FIRMPDEMP:PAYANN) %>%
  mutate(value = as.numeric(value)) %>% 
  select(GEOID, NAICS2017_LABEL, col_LABEL, name, value) %>% 
  pivot_wider(names_from = col_LABEL, values_from = value) 
}
```


Now, let's look at Asian business ownership in Arlington County, VA (`51013`)

```{r}
stco_abs_naics %>% 
  summarise_abs("51013",RACE_GROUP, c("60","96"))%>% 
  mutate(pct_asian = Asian/Classifiable) %>% 
  knitr::kable()
```

Or, Hispanic business ownership in professional services sector in Austin-Round Rock, TX metro area (`12420`)

```{r}
cbsa_abs_naics %>% 
  summarise_abs("12420",ETH_GROUP, c("020","096"), "54") %>% 
  mutate(pct_hispanic = Hispanic/Classifiable) %>% 
  knitr::kable()
```

